A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-10-05 paper.tex
minor edits

Of course, a space-efficient cast representation must be proved
correct. That is, a language implementation that uses space-efficient
casts should give the same output as the language semantics. Our
generic CEK machine comes with generic theorems mechanized in Agda
(Section~\ref{sec:framework}) that researchers can use to prove a cast
representation correct.  We believe this framework can reduce the
burden of proving correctness theorems and help people focus on the
cast representations. As a demonstration, we apply the framework to
Lazy UD coercions in normal form~\citep{Siek:2015:BCT:2737924.2737968}
and Lazy UD threesomes~\citep{Siek:2010:TWB:1706299.1706342}. We
believe the framework is also applicable to other cast
representations. The framework and demonstrations are available at the
following URL:
\[\text{\urlFrameworkAndIllustration}\]

@@

% All cast representation shall be proved correct, that is, besides reducing 
% space consumption, using a cast representation should not affect the results of 
% program execution.
In this section we present a framework for proving the 
correctness of space-efficient cast representations.
The framework includes an abstract data 
type named Cast ADT, which specifies the operators of cast representations, 
a CEK machine $\SMachine{C}$ parameterized over a Cast ADT,
which uses the cast operators implemented by $C$ to work with casts, 
and theorems relating $\SMachine{C}$ to $\CMachine{B}$.
%
Subsection~\ref{sec:cast-adt-and-space-efficient-cek} describes Cast ADT and the 
machine \SMachine{C}. We conjecture that the machine \SMachine{C} is 
space-efficient provided that the cast representation $C$ performs compression.
%
Subsection~\ref{sec:good-cast-adt} defines two subset of Cast ADT, namely Lazy D 
Cast ADT and Lazy UD Cast ADT. As their names suggest, if $C$ is a Lazy D Cast 
ADT, executing a program in $\SMachine{C}$ would give the same result as in 
$\LDMachine$. Similarly, if $C$ is a Lazy UD Cast ADT, executing a program in 
$\SMachine{C}$ would give the same result as in $\LUDMachine$.
%
We prove the theorems relating \SMachine{C} and \CMachine{B} in
Subsection~\ref{sec:S-correct}.
The framework is written in the Agda language~\citep{norell2008dependently}.

@@

\figref{fig:SEMachine-state} defines the syntax of the space-efficient CEK 
machine \SMachine{C}.
The key differences between \SMachine{C} and \CMachine{B} are in the definitions
of values and continuations, that is, places where casts can accumulate.

Let $u,v$ range over values.
%
Recall that in \CMachine{B} we have two value constructors for functions and 
another two for pairs. That setup allows zero-or-more layers of casts at the 
top of higher-order values, which might consume unbounded space. 
%
\SMachine{C} instead has exactly one layer of cast.
Thus non-casted values in \CMachine{B} correspond to \SMachine{C} values
where casts are identities ($\mathit{id}(T)$). The 
$\mathit{id}$ functions is provided by the Cast ADT.

Let $K$ ranges over continuations and $k$ ranges over pre-continuations.
This setup is essentially saying that cast frames and non-cast frames should
appear alternatively in continuations, and that a continuation always start
with a cast frame.
% %
% Pre-continuations are like the continuations in \CMachine{B}, but their 
% top-most frame must not be a cast frames.
% %
% A continuation is now a pre-continuation prefixed with a cast frame.
% %
% Recall that continuations in \CMachine{B} have zero or more casts at the top.
% In \SMachine{C}, however, every continuation has exactly one cast at the 
% top.
% %
\CMachine{B} continuations that have no casts at the top correspond
to \SMachine{C} continuations whose casts are identities.
\CMachine{B} continuations that have many casts at the top correspond
to \SMachine{C} continuations in where those casts are composed by
$\compose{c}{c}$, another functions provided by the Cast ADT.
%
Injectable types, environments, states, and cast results are the same as in \CBMachine.

Value typing (\valuetyping{v}{T}), continuation typing 
(\casttyping{k}{S}{T} and \casttyping{K}{S}{T}), and environment typing 
(\envtyping{\Gamma}{E}) are defined in the obvious way. 
They rely on cast typing \casttyping{c}{S}{T}, which shall be provided by the 
Cast~ADT.

>>>>

Of course, a space-efficient cast representation must be proved
correct. That is, a language implementation that uses space-efficient
casts should give the same output as the language semantics. Our
generic CEK machine comes with generic theorems mechanized in
the Agda proof assistant~\citep{norell2008dependently}
(Section~\ref{sec:framework}) that researchers can use to prove a cast
representation correct.  We believe this framework can reduce the
burden of proving correctness theorems and help people focus on the
cast representations. As a demonstration, we apply the framework to
Lazy UD coercions in normal form~\citep{Siek:2015:BCT:2737924.2737968}
and Lazy UD threesomes~\citep{Siek:2010:TWB:1706299.1706342}. We
believe the framework is also applicable to other cast
representations. The framework and demonstrations are available at the
following URL:
\[\text{\urlFrameworkAndIllustration}\]

@@

% All cast representation shall be proved correct, that is, besides reducing 
% space consumption, using a cast representation should not affect the results of 
% program execution.
In this section we present a framework for proving the correctness of
space-efficient cast representations.  The framework includes an
abstract data type named Cast ADT, which specifies the operators
required of a cast representation, a CEK machine $\SMachine{C}$
parameterized over a Cast ADT $C$, and theorems relating
$\SMachine{C}$ to $\CMachine{B}$.
%
Subsection~\ref{sec:cast-adt-and-space-efficient-cek} defines the Cast
ADT and the machine \SMachine{C}. We conjecture that the machine
\SMachine{C} is space-efficient provided that the cast representation
$C$ performs compression.
%
Subsection~\ref{sec:good-cast-adt} defines two subsets of the Cast
ADT, namely the Lazy D Cast ADT and the Lazy UD Cast ADT. As their
names suggest, if $C$ is a Lazy D Cast ADT, executing a program in
$\SMachine{C}$ would give the same result as in
$\LDMachine$. Similarly, if $C$ is a Lazy UD Cast ADT, executing a
program in $\SMachine{C}$ would give the same result as in
$\LUDMachine$.
%
We prove the theorems relating \SMachine{C} and \CMachine{B} in
Subsection~\ref{sec:S-correct}.

@@

\figref{fig:SEMachine-state} defines the syntax of the space-efficient
CEK machine \SMachine{C}.  The key differences between \SMachine{C}
and \CMachine{B} are the places where casts can accumulate, that is,
in the definitions of values and continuations.

Let $u,v$ range over values.
%
Recall that in \CMachine{B} we have two value constructors for
functions and another two for pairs. That setup allows zero-or-more
layers of casts at the top of higher-order values, which can consume
unbounded space.
%
\SMachine{C} instead has exactly one layer of cast.
Thus non-casted values in \CMachine{B} correspond to \SMachine{C} values
where casts are identities ($\mathit{id}(T)$). The 
$\mathit{id}$ functions is provided by the Cast ADT.

Let $K$ ranges over continuations and $k$ ranges over
pre-continuations.  This setup requires cast frames and non-cast
frames to alternate in continuations, and that a continuation always
starts with a cast frame.
% %
% Pre-continuations are like the continuations in \CMachine{B}, but their 
% top-most frame must not be a cast frames.
% %
% A continuation is now a pre-continuation prefixed with a cast frame.
% %
% Recall that continuations in \CMachine{B} have zero or more casts at the top.
% In \SMachine{C}, however, every continuation has exactly one cast at the 
% top.
% %
A \CMachine{B} continuation that does not have a cast at the top
corresponds to a \SMachine{C} continuation whose casts are identities.
A \CMachine{B} continuation that has many casts at the top corresponds
to a \SMachine{C} continuation in which those casts are composed by
$\compose{c}{c}$, which is an operation provided by the Cast ADT.
%
Injectable types, environments, states, and cast results are the same
as in \CBMachine.

Value typing (\valuetyping{v}{T}), continuation typing
(\casttyping{k}{S}{T} and \casttyping{K}{S}{T}), and environment
typing (\envtyping{\Gamma}{E}) are defined in the obvious way.  They
rely on cast typing \casttyping{c}{S}{T}, which is provided by the
Cast~ADT.
